Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Multi-Notch Searches

### 2017-07-12 paper.tex
eyreur

A 10ppm precursor mass tolerance is used for the initial calibration step, and then reduced to 5ppm for subsequent searches.
We allowed 120 modifications including PTMs, adducts, and chemical modifications in the database augmentation step.

@@

A critical parameter for peptide identification is mass accuracy\citep{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories: External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\citep{Olsen_2005}); and spectral calibration subsequent to the MS experiment (post-acquisition spectral calibration).
We developed a post-acquisition calibration procedure that builds upon the software lock mass concept\citep{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 0.01 Dalton fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired false discovery rate (e.g 1\%; based upon the target:decoy strategy\citep{Elias_2007}).
Step 2: The identifications from Step 1 are used to extract peak matches from the spectra, and a peak shifting procedure is performed.
The calibration algorithm is described in detail in Supplementary Material 1.

Calibration quality is evaluated through the results of a standard 10ppm precursor tolerance database search.
PSMs within 1\% FDR prior to calibration are compared with PSMs after the calibration procedure (Table~\ref{tbl:calib}).
The calibration procedure successfully addresses instrument miscalibration by centering the average error around zero.
The systemic noise introduced by various other factors is also addressed: this is evident in the decrease in the standard deviation of the errors in the precursor mass estimates, which show that the calibrated mass spectra have sub-ppm mass errors.
Histograms of the mass errors before and after calibration are plotted in Figure~\ref{fgr:figure1}.

\begin{table}[]
\centering
\caption{PSMs Within 1\% FDR Before and After Calibration}
\label{tbl:calib}
\begin{tabular}{llllll}

                                                                                &        & Orig        & Calib       & Orig        & Calib        \\
 &PSM Count                                                           &    152208       &        155617 &         217612   &   219172        \\
 &Average Error (ppm)                                                           &    -1.09         &       -0.16      &      -2.10      &        -0.16      \\
 &St. Dev of Error                                                     &     3.35        &      0.90       &       2.15      &        0.91      \\
Confidence Intervals& 68.3\% &   [-4.44,2.27]          &    [-1.06,0.74]          &    [-4.26,0.05]          &  [-1.08,0.75]            \\
                                                                                & 95.4\% &   [-7.80,5.62]          &    [-1.97,1.64]         &   [-6.41,2.20]          &      [-1.99,1.61]         \\
                                                                                & 99.7\% &   [-11.15,8.98]          &      [-2.87,2.54]        &    [-8.56,4.36]          &      [-2.90,2.58]         \\
\end{tabular}
\end{table}

%\begin{figure}
%  \includegraphics{calib1.png}
 % \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
 % \label{fgr:figure1}
%\end{figure}

@@

Prior to discussing the impact that multi-notch searches have on the G-PTM-D performance, it is useful to describe the variety of search tolerances used.
Traditional searches in bottom-up proteomics score potential matches between theoretical peptides and experimental spectra if the candidate peptide mass and the observed precursor mass are equal, within some defined tolerance (e.g. 5ppm, or 2.1 Da).
We'll refer to this type of search here as a \textit{single-notch} search.
The other common search mode, used in the original G-PTM-D publication, is a \textit{wide-parent tolerance search} ($\pm 500$ Da).
In that search, the precursor mass and the mass of the best matching theoretical peptide could be many Da apart.
An even less restrictive \textit{open-mass} search, matches peptides from a database to fragmentation spectra without taking the precursor mass into account at all.
We describe an alternative to these search modes: the multi-notch search.
The \textit{multi-notch} search is an extension of the single-notch search to multiple specific mass differences between precursor mass and theoretical mass~\ref{fig:fig1-searchtypes} where the mass difference is easily interpretable.
It can also be seen as a narrowing of the wide-parent tolerance search to only include prespecified mass differences.
The advantages and disadvantages of this alternative strategy are described below.

%\begin{figure}
%\includegraphics{fig1-searchTypes.png}
%\caption{Search Modes}
%\label{fig:fig1-searchtypes}
%\end{figure}

>>>>

A 10ppm precursor mass tolerance is used for the initial calibration step, and then reduced to 5ppm for subsequent searches.
We allowed 120 modifications including PTMs, adducts, and chemical modifications in the database augmentation step, see Supplementary Material 1 for a complete listing.

@@

A critical parameter for peptide identification is mass accuracy\citep{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories: External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\citep{Olsen_2005}); and spectral calibration subsequent to the MS experiment (post-acquisition spectral calibration).
We developed a post-acquisition calibration procedure that builds upon the software lock mass concept\citep{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 0.01 Dalton fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired false discovery rate (e.g 1\%; based upon the target:decoy strategy\citep{Elias_2007}).
Step 2: The identifications from Step 1 are used to extract peak matches from the spectra, and a peak shifting procedure is performed.
The calibration algorithm is described in detail in Supplementary Material 2.

We demonstrated the effectiveness of the calibration approach using a single CAST Mouse raw file as an example by displaying the errors in m/z values of MS1 peaks identified by Step 1 of the calibration, before and after the calibration procedure is applied.
In Figure [fgr:fig1-calibErrors] these errors are plotted versus two important variables: the m/z value of the peak and the retention time.
It can be seen in the graphs of Figure ### that calibration successfully addresses errors in non-linearity as a function of both retention time and m/z.
We note that in addition to these two parameters, the calibration function native to MetaMorpheus also considers non-linearities in peak intensity, injection time, total ion count, etc.
In addition, MetaMorpheus calibrates both MS1 and MS2 (using MS2 fragment matches), but only results from MS1 are shown here.
The complete list of parameters analyzed in the calibration is provided as Table $$$ in the supplement.

\begin{figure}[H]
 \includegraphics{fig1-calibErrors.png}
 \caption{Errors in m/z plotted vs the m/z value and the retention time}
 \label{fgr:fig1-calibErrors}
\end{figure}

Calibration quality is further evaluated through the results of a standard 10ppm precursor tolerance database search.
PSMs within 1\% FDR prior to calibration are compared with PSMs after the calibration procedure (Table~\ref{tbl:calib}).
The calibration procedure successfully addresses instrument miscalibration by centering the average error around zero.
The systemic noise introduced by various other factors is also addressed: this is evident in the decrease in the standard deviation of the errors in the precursor mass estimates, which show that the calibrated mass spectra have sub-ppm mass errors.
Histograms of the mass errors in identified peptides before and after calibration are plotted in Figure~\ref{fgr:fig2-10ppmSearchCalib}.

\begin{table}[]
\centering
\caption{PSMs Within 1\% FDR Before and After Calibration}
\label{tbl:calib}
\begin{tabular}{llll}

                                                                                &      & Orig        & Calib        \\
 &PSM Count                                                           &      217612   &   219172        \\
 &Average Error (ppm)                                                              &      -2.10      &        -0.16      \\
 &St. Dev of Error                                                    &       2.15      &        0.91      \\
Confidence Intervals& 68.3\%     &    [-4.26,0.05]          &  [-1.08,0.75]            \\
                                                                                & 95.4\%      &   [-6.41,2.20]          &      [-1.99,1.61]         \\
                                                                                & 99.7\%       &    [-8.56,4.36]          &      [-2.90,2.58]         \\
\end{tabular}
\end{table}

\begin{figure}[H]
 \includegraphics{fig2-10ppmSearchCalib.png}
 \caption{Errors in m/z plotted vs the m/z value and the retention time}
 \label{fgr:fig2-10ppmSearchCalib}
\end{figure}

@@

Prior to discussing the impact that multi-notch searches have on the G-PTM-D performance, it is useful to describe the variety of search tolerances used.
Traditional searches in bottom-up proteomics score potential matches between theoretical peptides and experimental spectra if the candidate peptide mass and the observed precursor mass are equal, within some defined tolerance (e.g. 5ppm, or 2.1 Da).
We'll refer to this type of search here as a \textit{single-notch} search.
The other common search mode, used in the original G-PTM-D publication, is a \textit{wide-parent tolerance search} ($\pm 500$ Da).
In that search, the precursor mass and the mass of the best matching theoretical peptide could be many Da apart.
An even less restrictive \textit{open-mass} search, matches peptides from a database to fragmentation spectra without taking the precursor mass into account at all.
We describe an alternative to these search modes: the multi-notch search.
The \textit{multi-notch} search is an extension of the single-notch search to multiple specific mass differences between precursor mass and theoretical mass~\ref{fig:fig3-searchTypes} where the mass difference is easily interpretable.
It can also be seen as a narrowing of the wide-parent tolerance search to only include prespecified mass differences.
The advantages and disadvantages of this alternative strategy are described below.

\begin{figure}[H]
\includegraphics{fig3-searchTypes.png}
\caption{Search Modes}
\label{fig:fig3-searchTypes}
\end{figure}
